---
output:
  github_document:
    fig_width: 5
    fig_height: 3
    df_print: kable
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE,
  message=FALSE,
  comment = "#>",
  fig.height=7,
  fig.width=8,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(knitr.table.format='markdown')
```

# venndir

<!-- badges: start -->
<!-- badges: end -->

The `venndir` package provides Venn directional
diagrams, that can optionally display item labels
inside the overlap regions.

The `pkgdown` reference: [jmw86069.github.io/venndir](https://jmw86069.github.io/venndir)


## Installation

The development version of venndir can be installed with:

``` r
# install.packages("remotes")
remotes::install_github("jmw86069/venndir");
```

## Why Venndir?

Biological data often involves components, for example genes, and
the direction of change, for example increase or decrease. It is not
enough to compare genes affected by two experiments, it is critical
to compare the direction of change. How else can we discriminate
the effects of disease from the effects of a cure?

```{r venn_intro}
# silence the warnings
options("warn"=-1)

library(venndir)
vo <- venndir(make_venn_test(100, 2, do_signed=TRUE),
   proportional=TRUE,
   overlap_type="each", font_cex=1.2)

```

## Features of venndir

The core is `venndir()` which takes a `setlist` as
input and produces a Venn diagram. When the `setlist`
contains directionality (sign), the directional
overlaps are also displayed.

To illustrate the point, `make_venn_test()` is used
to create test `setlist` data.

```{r venn_setlist}
setlist <- make_venn_test(100, 3)
setlist
```

A `setlist` is a list of vectors. The names of the list,
one for each vector, are the set names. Each vector contains
items which are the subject of the Venn overlaps.

Given a `setlist`, you can create a Venn diagram with `venndir()`:

```{r venn_1}
venndir(setlist)
```

You can make a proportional Venn diagram,
also known as a Euler diagram. Add argument `proportional=TRUE`.

```{r venn_1e}
vo <- venndir(setlist,
   proportional=TRUE)
```


## Venn Direction

The namesake of this package is Venn with directionality!

Test with: `make_venn_test(..., do_signed=TRUE)`.
Each item is associated with a numerical direction:

* `+1` for up
* `-1` for down

Each vector in `setlist_dir` is a named vector,
whose **names** are the items, and whose **values** are
the direction, with `+1` or `-1`. Take a look.

```{r venn_setlist_signed}
setlist_dir <- make_venn_test(100, 3, do_signed=TRUE)
setlist_dir
```


For biological data, direction is important and relevant.
Whether a gene is regulated up or down might be the difference
between disease and treatment.

> Note `make_venn_tests()` can simulate concordance,
and the default is `concordance=0.5`. Concordance is a
measure of how frequently two directions are the same, and
is defined `(agree - disagree) / (n)`. Thus, `concordance=0` means
there are the same number that agree as disagree in direction,
and `concordance=1` means every element agrees in direction.

There are a few ways to summarize directional overlaps,
which vary by the amount of detail.

* `overlap_type="overlap"` - This method ignores direction.
* `overlap_type="agreement"` - This method displays agreement
and disagreement, with no details about up/down direction.
* `overlap_type="concordance"` - **(default)** This method displays
concordant directions, up-up, down-down; all discordant
combinations are grouped together as "discordant".
* `overlap_type="each"` - This method displays each directional
overlap.


### overlap_type="concordance"

This option displays the number `up-up`, and the number `down-down`,
and everything else is considered  "discordant".
This approach is effective at conveying direction, without too many
details.

Notice the `"X"` to indicate discordance.

```{r venndir_1}
venndir(setlist_dir)
```


### overlap_type="each"

This option shows the count for each combination.
It works best when you want to see all the details, however
it can create a lot of labels!

```{r venndir_each}
venndir(setlist_dir, overlap_type="each")
```


### overlap_type="agreement"

This option shows the count that agrees in direction, and
the count that disagrees in direction. It does not indicate
whether the agreement is up or down.

This option is especially good at summarizing the
number that agree and disagree, without including 
potentially confusing details.

```{r venndir_agreement}
venndir(setlist_dir, overlap_type="agreement")
```

### overlap_type="overlap"

This option only displays the overlap count, ignoring direction.

```{r venndir_overlap}
venndir(setlist_dir, overlap_type="overlap")
```


## Proportional Venn Direction

As mentioned previously, you can display proportional Venn diagram,
(a Euler diagram), which uses the excellent `eulerr` R package.
Add argument `proportional=TRUE`.

```{r venndir_each_p}
vo <- venndir(setlist_dir,
   proportional=TRUE,
   font_cex=c(1.3, 0.9, 0.7))
```

Labeling is often a challenge with proportional Venn diagrams.
For very small regions, you can push the label outside
with `inside_percent_threshold=5`. This option will
move labels outside when the region is less than 5% of
the total area.

```{r venndir_overlap_p}
venndir(setlist,
   proportional=TRUE,
   overlap_type="overlap",
   label_style="lite box",
   inside_percent_threshold=5,
   font_cex=c(1.3, 1))
```


## Customizing the Venn diagram

### Label styles

The argument `label_style` can be used to customize the label:

* `label_style="lite"` - adds lite shading behind each label
* `label_style="shaded"` - adds partially transparent shading
* `label_style="fill"` - adds solid colored shading
* `label_style="box"` - adds a small outline box around the label

Multiple terms can be combined, for example to add shading and a box:

* `label_style="shaded box"` - adds colored shading and a box outline

### Label position

Argument `show_labels` is used to position labels.
Each letter defines a type of label, and UPPERCASE or lowercase
indicates where to place the label.

* UPPERCASE = outside the Venn diagram
* lowercase = inside the Venn diagram

The letters:

* N = the set name
* C = the overlap count
* S = the signed overlap count(s)
* i = the overlapping items

Guidance:

* The default: `show_labels="Ncs"` will show _N_ame outside, _c_ounts inside.
When _s_igned labels are shown, they also appear inside.
Signed labels are not shown when `overlap_type="overlap"`.
* Display all labels inside: `show_labels="ncs"`
* It works best to have the _c_ounts and _s_igned counts together,
usually inside.
* To display items, `show_labels="Ni"` is recommended, to show _N_ame
outside, and _i_tems inside. You can still use `show_items="sign item"`
so that each item label will include the direction.
* When displaying items, the counts and signed counts are automatically
moved outside. (There isn't a great way to place item labels around
the count labels. Maybe in future.)


```{r label_preset_1}
vo4 <- venndir(setlist,
   show_labels="ncs",
   inside_percent_threshold=0)
```

Hide line segments with `show_segments=FALSE`

```{r label_preset_1l}
vo4l <- venndir(setlist,
   show_labels="Ncs",
   show_segments=FALSE,
   inside_percent_threshold=0)
```


## Text Venn for the R Console

There is a text Venn diagram, surprisingly useful for quickly
checking overlaps and direction. Note that the R console,
and R help examples display colored text, just not in
Rmarkdown.

The first example is the basic Venn overlap, without direction.

```{r textvenn_1}
# Options are used for the RMarkdown
# options("jam.htmlOut"=TRUE, "jam.comment"=FALSE)

setlist <- make_venn_test(1000, 3, do_signed=TRUE)
names(setlist) <- gsub("set_", "", names(setlist));
textvenn(setlist, overlap_type="overlap")
```

But of course direction is helpful, so here it is with the
default `overlap_type="concordance"` (below)

```{r textvenn_2}
textvenn(setlist, overlap_type="concordance")
```

Not all consoles can display Unicode arrows, so you can use ASCII
output only with `unicode=FALSE`:

```{r textvenn_3}
textvenn(setlist, overlap_type="concordance", unicode=FALSE)

# Revert options
# options("jam.htmlOut"=FALSE, "jam.comment"=TRUE)
```

Sorry, no proportional text Venn diagrams (yet)!


## Nudge Venn circles

Another driving reason for this package is that sometimes
proportional Venn (Euler) diagrams fail to produce circles
that show all the overlaps. While sometimes it is
mathematically impossible, other times are just... puzzling.

See below, the overlap `set_A&set_B` has 1 count, but is not
displayed.

```{r nudge_1}
overlaps <- c(set_A=187, set_B=146, set_C=499,
   `set_A&set_B`=1,
   `set_A&set_C`=181,
   `set_B&set_C`=219,
   `set_A&set_B&set_C`=20);
# convert to setlist
setlist_o <- counts2setlist(overlaps)

vn <- venndir(setlist_o,
   expand_fraction=0.15,
   proportional=TRUE,
   font_cex=1.4,
   set_colors=c("firebrick2", "dodgerblue", "#9999AA"))
```

The argument `circle_nudge` lets you nudge (move) a Venn circle
given x,y coordinates. Provide a `list` named by the set
you want to move, with a `numeric` vector for the `x,y`
coordinates direction.

```{r nudge_2}
vo_nudge <- venndir(setlist_o,
   expand_fraction=0.15,
   # label_style="lite box",
   font_cex=1.4,
   proportional=TRUE,
   circle_nudge=list(set_A=c(1, 0), set_B=c(-1, 0)),
   set_colors=c("firebrick2", "dodgerblue", "#9999AA"))
```


## Item labels

An optional but useful feature is to include item labels
inside the Venn diagram. It helps answer the question,
"What are those shared items?" (In my experience, that's
a very early question.)

This step can also include the
directional sign, which helps indicate which items
are shared, and whether they have the same direction.
In cases with too many labels to display, it is sometimes
still useful to display the sign, as a visual cue for the
proportion of shared or discordant signs.

Two changes are required:

1. Argument `show_labels` must include `"i"` to indicate items should
be included in the labels. Preferred options:

   * `"Ni"` which shows _N_ame outside, and _i_tems inside.
   * `"NCi"` which shows _N_ame and _C_ounts outside, and _i_tems inside.

2. Argument `show_items` should be one of

   * `"item"` - to show only the item label
   * `"sign item"` - to show the sign beside each item label
   * `"sign"` - to show only the item label

```{r vennitems_1}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
   poly_alpha=0.3,
   show_labels="NCi",
   show_items="sign item");
```

Interestingly, the density of labels gives some indication
of the relative overlaps.

The same plot using proportional circles makes the
label density effectively uniform.
Note the option `show_items="sign"` displays only the directional arrow,
and `item_cex=2` makes the arrows twice as large as normal.


```{r vennitems_1p}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
   poly_alpha=0.3,
   show_labels="Ni",
   item_cex=2,
   show_items="sign",
   proportional=TRUE);
```

The sign is an interesting visual summary when there are too
many labels to display otherwise.

```{r vennitems_2}
setlist <- make_venn_test(1000, 3, do_signed=TRUE);
venndir(setlist,
   show_labels="Ni",
   show_items="sign",
   item_cex=2,
   expand_fraction=0.1,
   show_segments=FALSE,
   max_items=10000);
```

Again, proportional Venn circles effectively makes the density
uniform.

```{r vennitems_2p}
venndir(setlist,
   show_labels="Ni",
   overlap_type="each",
   show_items="sign",
   item_cex=2,
   expand_fraction=0.1,
   max_items=10000,
   show_segments=FALSE,
   proportional=TRUE)
```
