---
output:
  github_document:
    fig_width: 5
    fig_height: 3
    df_print: kable
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE,
  message=FALSE,
  comment = "#>",
  fig.height=8,
  fig.width=8,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(knitr.table.format='markdown')
```

# venndir

<!-- badges: start -->
<!-- badges: end -->

The goal of venndir is to enable directional Venn overlap
analysis and visualization.


## Installation

The development version of venndir can be installed with:

``` r
# install.packages("remotes")
remotes::install_github("jmw86069/venndir");
```


## Brief Overview of venndir

The `venndir` package includes a utility function `make_venn_test()`
to create test data.

```{r venn_1}
library(venndir)
setlist <- make_venn_test(100, 3)
venndir(setlist)
```

You can make a proportional Venn diagram,
also known as a Euler diagram. More examples of proportional
Venn diagrams are described below, with custom options to manipulate
the position of circles and labels.

```{r venn_1e}
venndir(setlist, proportional=TRUE)
```

The default output is a base R plot, but you can use `ggplot2`
with argument `plot_style="gg"`. The function returns the `ggplot`
object which can be manipulated alongside other `grid`
graphical objects.

```{r venn_1g}
venndir(setlist, plot_style="gg")
```

## Venn Direction

For a more interesting case, `make_venn_test()` can make
sets where the items have directionality, encoded with `1` for up,
or `-1` for down.

For biological data, direction is important and relevant.
Whether a gene is regulated up or down might be the difference
between disease and treatment.

> Note `make_venn_tests()` can simulate concordance,
and the default is `concordance=0.5`. Concordance is a
measure of how frequently two directions are the same, and
is defined `(agree - disagree) / (n)`. Thus, `concordance=0` means
there are the same number that agree as disagree in direction,
and `concordance=1` means every element agrees in direction.

There are a few ways to think about "concordance" -- the measure
of agreement in direction, and so there are few ways to display
concordance inside a Venn diagram.
The default argument
`overlap_type="concordance"` displays
the number `up-up`, and the number `down-down`, and everything else
is considered  "discordant".

```{r venndir_1}
setlist <- make_venn_test(1000, 3, do_signed=TRUE)
venndir(setlist)
```

This style is effective for 3 or more sets, where the number
of possible discordant combinations may be too many to display
comfortably. In this case 
it is still useful to see how many agree up, and how many agree down,
the rest simply disagree.


Another option `overlap_type="each"` shows the count for
every combination. As said before, it works best for 2 or 3 sets.

```{r venndir_each}
venndir(setlist, overlap_type="each")
```


The option `overlap_type="agreement"` shows only
"agreement" and "mixed", regardless of the direction.
This option is especially good at summarizing the
number that agree and disagree.

```{r venndir_agreement}
venndir(setlist, overlap_type="agreement")
```

The last option simply turns off the display of
direction, `overlap_type="overlap"`.

```{r venndir_overlap}
venndir(setlist, overlap_type="overlap")
```



## Proportional Venn Direction

As shown above, you can switch output to 
a proportional Venn diagram, which uses the
really nice `eulerr` R package.

```{r venndir_each_p}
venndir(setlist, proportional=TRUE, distance=5, label_style="lite",
   font_cex=c(1.3, 1))
```

Labeling is often a challenge with proportional Venn
diagrams. By default, for directional Venn counts,
if a polygon is less than 5% of the total area,
the label is placed outside. For standard Venn
counts the default threshold is 1%.

Below is the same diagram, showing only the overlap
counts, and not the directional counts.

```{r venndir_overlap_p}
venndir(setlist,
   proportional=TRUE,
   overlap_type="overlap",
   label_style="lite",
   font_cex=c(1.3, 1))
```

You can change the threshold for placing labels
outside with the argument `inside_percent_threshold`,
which takes integer percent values from 0 to 100.

```{r venndir_overlap_p2}
venndir(setlist,
   proportional=TRUE,
   inside_percent_threshold=0.1,
   overlap_type="overlap",
   font_cex=c(1.3, 1))
```

## Customizing the Venn diagram

The output of `venndir()` is a `list` with:

* `"venn_spdf"` - which contains polygon coordinates
stored as a `sp::SpatialPolygonsDataFrame`. Essentially
it stores each Venn polygon, with annotations in a `data.frame`.
* `"label_df"` - a `data.frame` with individual text labels,
colors, and coordinates.

You can edit the `label_df` data manually, as needed, then
render the Venn diagram using `render_venndir()`.
(I really like having a function named `render_venndir()`...)

```{r venndir_each_p2}
setlist <- make_venn_test(1000, 3, do_signed=FALSE)
vo <- venndir(setlist, proportional=TRUE, do_plot=FALSE);

print(head(vo$label_df));

vo$label_df[1:3,"border"] <- c("red4", "darkorange", "blue4");
vo$label_df[1:3,"fill"] <- c("red3", "darkorange2", "blue3");
render_venndir(vo);
```


A convenience function `venndir_label_style()` is provided
which provides two features:

1. It can adjust a label positions and visibility
using `label_preset`.
2. It can adjust label visual style using `label_style`.

### venndir_label_style() and label_preset

The `label_preset` has a few pre-configured options:

* `"main inside"` - displays main set labels, and counts
inside each polygon.
* `"main outside"` - displays each set label outside,
and counts inside.
* `"outside"` - displays set labels and counts outside.
* `"main items"` - displays set names outside, and item labels
inside each polygon. See below for examples.

```{r label_preset_1}
vo4 <- venndir_label_style(vo,
   label_preset="main outside",
   inside_percent_threshold=0)
render_venndir(vo4);
```

You can hide line segments with `show_segments=FALSE`.

```{r label_preset_1l}
vo4l <- venndir_label_style(vo,
   label_preset="main outside",
   inside_percent_threshold=0)
render_venndir(vo4l,
   show_segments=FALSE);
```



### venndir_label_style() and label_style

The `label_style` is used for visual effects, to improve
visibility of the text labels. It applies two basic operations,
fill and border.

Fill options:

* `"basic"` - removes background fill
* `"shaded"` - partial transparent fill using the overlap color
* `"fill"` - fill using the overlap color
* `"lite"` - lite shaded fill
* `"custom"` - will not update the fill, in case you manually adjusted these values

Border options:

* `"box"` - will draw a border around each label
*`* `""` - absence of `"box"` in `label_style` will remove any border

The `label_style` string can be any string that contains those
values, for example:

* `label_style="lite box"`
* `label_style="shaded"`
* `label_style="basic box"`

```{r label_style_1}
vo3 <- venndir_label_style(vo,
   label_style="lite box")
render_venndir(vo3);
```


## Text Venn for the R Console

There is a text Venn diagram, surprisingly useful for quickly
checking overlaps and direction. Note that the R console,
and R help examples display colored text, just not in
Rmarkdown.

The first example is the basic Venn overlap, without direction.

```{r textvenn_1}
setlist <- make_venn_test(1000, 3, do_signed=TRUE)
textvenn(setlist, overlap_type="overlap")
```

But of course direction is helpful, so here it is with the
default `overlap_type="concordance"` (below)

```{r textvenn_2}
textvenn(setlist, overlap_type="concordance")
```

Not all consoles can display Unicode arrows, so you can use ASCII
output only with `unicode=FALSE`:

```{r textvenn_3}
textvenn(setlist, overlap_type="concordance", unicode=FALSE)
```

Sorry, no proportional text Venn diagrams (yet)!


## Nudge Venn circles

Another driving reason for this package is that sometimes
proportional Venn (Euler) diagrams fail to produce circles
that show all the overlaps. While sometimes it is
mathematically impossible, other times are just... puzzling.

See below, the overlap `set_A&set_B` has 1 count, but is not
displayed.

```{r nudge_1}
overlaps <- c(set_A=187, set_B=146, set_C=499,
   `set_A&set_B`=1,
   `set_A&set_C`=181,
   `set_B&set_C`=219,
   `set_A&set_B&set_C`=20);
# convert to setlist
setlist_o <- counts2setlist(overlaps)

venndir(setlist_o,
   proportional=TRUE,
   font_cex=1.4,
   set_colors=c("firebrick2", "dodgerblue", "#9999AA"))
```

The argument `circle_nudge` lets you nudge (move) a Venn circle
given x,y coordinates. Provide a `list` named by the set
you want to move, with a `numeric` vector for the `x,y`
coordinates direction.

```{r nudge_2}
vo_nudge <- venndir(setlist_o,
   font_cex=1.4,
   proportional=TRUE,
   circle_nudge=list(set_A=c(1, 0), set_B=c(-1, 0)),
   set_colors=c("firebrick2", "dodgerblue", "#9999AA"))
```


## Item labels

An optional but useful feature is to include item labels
inside the Venn diagram. It helps answer the question,
"What are those shared items?" (In my experience, that's
a very early question.)

This step can also include the
directional sign, which helps indicate which items
are shared, and whether they have the same direction.
In cases with too many labels to display, it is sometimes
still useful to display the sign, as a visual cue for the
proportion of shared or discordant signs.

Two arguments are required: the `label_preset` to define the
visibility of items and main set labels; and `show_items` to
define the type of item label.
For example `label_preset="main items"` will display main set
labels outside, and item labels inside;
and `show_items="item"` will display the item label,
`show_items="sign item"` will display the sign and item label,
`show_items="sign"` will display only the sign.

```{r vennitems_1}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
   poly_alpha=0.3,
   label_preset="main items",
   show_items="sign item");
```

Interestingly, the density of labels gives some indication
of the relative overlaps.

The same plot using proportional circles makes the
label density effectively uniform:


```{r vennitems_1p}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
   poly_alpha=0.3,
   label_preset="main items",
   show_items="item",
   proportional=TRUE);
```

With too many features to label, it's still interesting to indicate
the sign.

```{r vennitems_2}
setlist <- make_venn_test(1000, 3, do_signed=TRUE);
venndir(setlist,
   label_preset="main items",
   show_items="sign",
   show_segments=FALSE,
   max_items=10000);
```

Again, proportional Venn circles effectively makes the density
uniform.

```{r vennitems_2p}
venndir(setlist,
   label_preset="main items",
   overlap_type="each",
   show_items="sign",
   max_items=10000,
   show_segments=FALSE,
   proportional=TRUE);
```
