---
output:
  github_document:
    fig_width: 5
    fig_height: 3
    df_print: kable
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE,
  message=FALSE,
  comment = "#>",
  fig.height=8,
  fig.width=8,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(knitr.table.format='markdown')
```

# venndir

<!-- badges: start -->
<!-- badges: end -->

The `venndir` package provides Venn directional
diagrams, that can optionally display item labels
inside the overlap regions.

The `pkgdown` reference: [jmw86069.github.io/venndir](https://jmw86069.github.io/venndir)


## Installation

The development version of venndir can be installed with:

``` r
# install.packages("remotes")
remotes::install_github("jmw86069/venndir");
```


## Features of venndir

The core is `venndir()` which takes a `setlist` as
input and produces a Venn diagram. When the `setlist`
contains directionality (sign), the directional
overlaps are also displayed.

To illustrate the point, `make_venn_test()` is used
to create test `setlist` data.

```{r venn_setlist}
library(venndir)
setlist <- make_venn_test(100, 3)
setlist
```

A `setlist` is a list of vectors. The names of the list,
one for each vector, are the set names. Each vector contains
items which are the subject of the Venn overlaps.

Given a `setlist`, you can create a Venn diagram with `venndir()`:

```{r venn_1}
venndir(setlist)
```

You can make a proportional Venn diagram,
also known as a Euler diagram. More examples of proportional
Venn diagrams are described below, but for now the
simplest approach is to add argument `proportional=TRUE`:

```{r venn_1e}
venndir(setlist, proportional=TRUE)
```

The default output use base R graphics, but you can
use `ggplot2` with the argument `plot_style="gg"`:

```{r venn_1g}
venndir(setlist, proportional=TRUE, plot_style="gg")
```

The function invisibly returns the `ggplot`
object which can be manipulated alongside other `grid`
graphical objects.

## Venn Direction

For a more interesting case, `make_venn_test(..., do_signed=TRUE)`
creates a `setlist` with directionality (sign), which means
each item is associated with a numerical direction:

* `+1` for up
* `-1` for down

In this case, each vector in the `setlist` is a named vector,
whose **names** are the items, and whose **values** are
the signed direction `+1` or `-1`. Take a look.

```{r venn_setlist_signed}
setlist_dir <- make_venn_test(100, 3, do_signed=TRUE)
setlist_dir
```


For biological data, direction is important and relevant.
Whether a gene is regulated up or down might be the difference
between disease and treatment.

> Note `make_venn_tests()` can simulate concordance,
and the default is `concordance=0.5`. Concordance is a
measure of how frequently two directions are the same, and
is defined `(agree - disagree) / (n)`. Thus, `concordance=0` means
there are the same number that agree as disagree in direction,
and `concordance=1` means every element agrees in direction.

There are a few ways to summarize directional overlaps,
which vary by the amount of detail.

* `overlap_type="overlap"` - This method ignores direction.
* `overlap_type="agreement"` - This method displays agreement
and disagreement, with no details about up/down direction.
* `overlap_type="concordance"` - **(default)** This method displays
concordant directions, up-up, down-down; all discordant
combinations are grouped together as "discordant".
* `overlap_type="each"` - This method displays each directional
overlap.


### overlap_type="concordance"

Let's start with `overlap_type="concordance"` which displays
the number `up-up`, and the number `down-down`, and everything else
is considered  "discordant". This approach is effective
at conveying direction, without too many unhelpful details.

```{r venndir_1}
venndir(setlist_dir)
```

It is mostly helpful when you want to see how many overlaps
agree up and down, and when the overlaps that disagree
in direction are less interesting. For example, sometimes
the overlaps are mostly `up-up` and rarely `down-down`,
which can be very helpful to know.


### overlap_type="each"

The option `overlap_type="each"` shows the count for
each directional combination. It works best when you
want to see all the details, typically for only
2 or 3 sets.

```{r venndir_each}
venndir(setlist_dir, overlap_type="each")
```


### overlap_type="agreement"

The option `overlap_type="agreement"` shows the count
for each overlap that agrees in direction, regardless
of the direction; and the count for overlaps that
disagree in direction.

This option is especially good at summarizing the
number that agree and disagree, without including 
potentially confusing details.

```{r venndir_agreement}
venndir(setlist_dir, overlap_type="agreement")
```

### overlap_type="overlap"

The option `overlap_type="overlap"` simply ignores
directionality.

```{r venndir_overlap}
venndir(setlist_dir, overlap_type="overlap")
```


## Proportional Venn Direction

As shown above, you can switch output to 
a proportional Venn diagram, which uses the
really nice `eulerr` R package.

```{r venndir_each_p}
venndir(setlist_dir,
   proportional=TRUE,
   #distance=5, label_style="lite",
   font_cex=c(1.3, 0.9))
```

Labeling is often a challenge with proportional Venn
diagrams. By default, for directional Venn counts,
if a polygon is less than 5% of the total area,
the label is placed outside. For standard Venn
counts the default threshold is 1%.

Below is the same diagram, showing only the overlap
counts, and not the directional counts.

```{r venndir_overlap_p}
venndir(setlist,
   proportional=TRUE,
   overlap_type="overlap",
   label_style="lite",
   font_cex=c(1.3, 1))
```

You can change the threshold for placing labels
outside with the argument `inside_percent_threshold`,
which takes integer percent values from 0 to 100.

```{r venndir_overlap_p2}
venndir(setlist,
   proportional=TRUE,
   inside_percent_threshold=0.1,
   overlap_type="overlap",
   font_cex=c(1.3, 1))
```

## Customizing the Venn diagram

The output of `venndir()` is a `list` with:

* `"venn_spdf"` - which contains polygon coordinates
stored as a `sp::SpatialPolygonsDataFrame`. Essentially
it stores each Venn polygon, with annotations in a `data.frame`.
* `"label_df"` - a `data.frame` with individual text labels,
colors, and coordinates.

You can edit the `label_df` data manually, as needed, then
render the Venn diagram using `render_venndir()`.
(I really like having a function named `render_venndir()`...)

```{r venndir_each_p2}
setlist <- make_venn_test(1000, 3, do_signed=FALSE)
vo <- venndir(setlist, proportional=TRUE, do_plot=FALSE);

print(head(vo$label_df));

vo$label_df[1:3,"border"] <- c("red4", "darkorange", "blue4");
vo$label_df[1:3,"fill"] <- c("red3", "darkorange2", "blue3");
render_venndir(vo);
```


A convenience function `venndir_label_style()` is provided
which provides two features:

1. It can adjust a label positions and visibility
using `label_preset`.
2. It can adjust label visual style using `label_style`.

### venndir_label_style() and label_preset

The `label_preset` has a few pre-configured options:

* `"main inside"` - displays main set labels, and counts
inside each polygon.
* `"main outside"` - displays each set label outside,
and counts inside.
* `"outside"` - displays set labels and counts outside.
* `"main items"` - displays set names outside, and item labels
inside each polygon. See below for examples.

```{r label_preset_1}
vo4 <- venndir_label_style(vo,
   label_preset="main outside",
   inside_percent_threshold=0)
render_venndir(vo4);
```

You can hide line segments with `show_segments=FALSE`.

```{r label_preset_1l}
vo4l <- venndir_label_style(vo,
   label_preset="main outside",
   inside_percent_threshold=0)
render_venndir(vo4l,
   show_segments=FALSE);
```



### venndir_label_style() and label_style

The `label_style` is used for visual effects, to improve
visibility of the text labels. It applies two basic operations,
fill and border.

Fill options:

* `"basic"` - removes background fill
* `"shaded"` - partial transparent fill using the overlap color
* `"fill"` - fill using the overlap color
* `"lite"` - lite shaded fill
* `"custom"` - will not update the fill, in case you manually adjusted these values

Border options:

* `"box"` - will draw a border around each label
*`* `""` - absence of `"box"` in `label_style` will remove any border

The `label_style` string can be any string that contains those
values, for example:

* `label_style="lite box"`
* `label_style="shaded"`
* `label_style="basic box"`

```{r label_style_1}
vo3 <- venndir_label_style(vo,
   label_style="lite box")
render_venndir(vo3);
```


## Text Venn for the R Console

There is a text Venn diagram, surprisingly useful for quickly
checking overlaps and direction. Note that the R console,
and R help examples display colored text, just not in
Rmarkdown.

The first example is the basic Venn overlap, without direction.

```{r textvenn_1}
setlist <- make_venn_test(1000, 3, do_signed=TRUE)
textvenn(setlist, overlap_type="overlap")
```

But of course direction is helpful, so here it is with the
default `overlap_type="concordance"` (below)

```{r textvenn_2}
textvenn(setlist, overlap_type="concordance")
```

Not all consoles can display Unicode arrows, so you can use ASCII
output only with `unicode=FALSE`:

```{r textvenn_3}
textvenn(setlist, overlap_type="concordance", unicode=FALSE)
```

Sorry, no proportional text Venn diagrams (yet)!


## Nudge Venn circles

Another driving reason for this package is that sometimes
proportional Venn (Euler) diagrams fail to produce circles
that show all the overlaps. While sometimes it is
mathematically impossible, other times are just... puzzling.

See below, the overlap `set_A&set_B` has 1 count, but is not
displayed.

```{r nudge_1}
overlaps <- c(set_A=187, set_B=146, set_C=499,
   `set_A&set_B`=1,
   `set_A&set_C`=181,
   `set_B&set_C`=219,
   `set_A&set_B&set_C`=20);
# convert to setlist
setlist_o <- counts2setlist(overlaps)

venndir(setlist_o,
   proportional=TRUE,
   font_cex=1.4,
   set_colors=c("firebrick2", "dodgerblue", "#9999AA"))
```

The argument `circle_nudge` lets you nudge (move) a Venn circle
given x,y coordinates. Provide a `list` named by the set
you want to move, with a `numeric` vector for the `x,y`
coordinates direction.

```{r nudge_2}
vo_nudge <- venndir(setlist_o,
   font_cex=1.4,
   proportional=TRUE,
   circle_nudge=list(set_A=c(1, 0), set_B=c(-1, 0)),
   set_colors=c("firebrick2", "dodgerblue", "#9999AA"))
```


## Item labels

An optional but useful feature is to include item labels
inside the Venn diagram. It helps answer the question,
"What are those shared items?" (In my experience, that's
a very early question.)

This step can also include the
directional sign, which helps indicate which items
are shared, and whether they have the same direction.
In cases with too many labels to display, it is sometimes
still useful to display the sign, as a visual cue for the
proportion of shared or discordant signs.

Two arguments are required: the `label_preset` to define the
visibility of items and main set labels; and `show_items` to
define the type of item label.
For example `label_preset="main items"` will display main set
labels outside, and item labels inside;
and `show_items="item"` will display the item label,
`show_items="sign item"` will display the sign and item label,
`show_items="sign"` will display only the sign.

```{r vennitems_1}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
   poly_alpha=0.3,
   label_preset="main items",
   show_items="sign item");
```

Interestingly, the density of labels gives some indication
of the relative overlaps.

The same plot using proportional circles makes the
label density effectively uniform:


```{r vennitems_1p}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
   poly_alpha=0.3,
   label_preset="main items",
   show_items="item",
   proportional=TRUE);
```

With too many features to label, it's still interesting to indicate
the sign.

```{r vennitems_2}
setlist <- make_venn_test(1000, 3, do_signed=TRUE);
venndir(setlist,
   label_preset="main items",
   show_items="sign",
   show_segments=FALSE,
   max_items=10000);
```

Again, proportional Venn circles effectively makes the density
uniform.

```{r vennitems_2p}
venndir(setlist,
   label_preset="main items",
   overlap_type="each",
   show_items="sign",
   max_items=10000,
   show_segments=FALSE,
   proportional=TRUE);
```
