---
output:
  github_document:
    fig_width: 5
    fig_height: 3
    df_print: kable
    dev: ragg_png
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE,
  message=FALSE,
  comment = "#>",
  fig.height=7,
  fig.width=8,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(knitr.table.format='markdown')
options(warn=FALSE)
```

# venndir

<!-- badges: start -->
<!-- badges: end -->

The `venndir` package provides Venn directional
diagrams, that can optionally display item labels
inside the overlap regions.

The `pkgdown` reference: [jmw86069.github.io/venndir](https://jmw86069.github.io/venndir)


## Installation

The development version of venndir can be installed with:

``` r
# install.packages("remotes")
remotes::install_github("jmw86069/venndir");
```

## Why Venndir?

Biological data often involves components, for example genes, and
the direction of change, for example increase or decrease. It is not
enough to compare genes affected by two experiments, it is critical
to compare the direction of change. How else can we discriminate
the effects of disease from the effects of a cure?

```{r venn_intro, fig.alt="Venn diagram with directional counts included"}
library(venndir)

vo <- venndir(make_venn_test(100, 2, do_signed=TRUE),
   proportional=TRUE,
   overlap_type="each", font_cex=1.2)
```

## Features of venndir

The core is `venndir()` which takes a `setlist` as
input and produces a Venn diagram. When the `setlist`
contains directionality (sign), the directional
overlaps are also displayed.

To illustrate the point, `make_venn_test()` is used
to create test `setlist` data.

```{r venn_setlist}
setlist <- make_venn_test(100, 3)
```

* `setlist` is a `list` of `character` vectors.
* The names of the list `names(setlist)` are the set names.
* Each vector contains items which are the subject of the Venn overlaps.

Now you can create a Venn diagram with `venndir()`:

```{r venn_1, fig.alt="Venn diagram showing default output. There are no directional counts because the setlist did not contain directional information."}
venndir(setlist)
```

You can make a proportional Venn diagram, known as a Euler diagram.
Add argument `proportional=TRUE`.

```{r venn_1e, fig.alt="Euler diagram showing a proportional Venn diagram."}
vo <- venndir(setlist,
   proportional=TRUE)
```


## Venn Direction

The namesake of this package is Venn with directionality! **VennDir**

Use the testing function to create example data:  
`make_venn_test(..., do_signed=TRUE)`

The `setlist` is a `list` as before, except now the values are:

* `+1` for up
* `-1` for down

Each vector is a named vector,  
whose **vector names** are the items, and  
whose **vector values** are the directions.  
Take a look.

```{r venn_setlist_signed}
setlist_dir <- make_venn_test(100, 3, do_signed=TRUE)
setlist_dir
```


**For biological data, direction is important, and relevant**.
Up- or down-regulation might be the difference between disease
and treatment.

> Aside: `make_venn_tests()` can simulate concordance.  
Concordance measures how frequently two directions are the same.
Kruskal's Concordance is defined: `(agree - disagree) / (n)`.  
Thus, `concordance=0` means there are the same number that agree
as disagree in direction, and `concordance=1` means every element
agrees in direction.

There are different ways to visualize directionality:

1. `overlap_type="concordance"` - **(default for signed input)**:
Concordant up `R cat("\u2191")`, down `R cat("\u2193")`, and discordant X.
2. `overlap_type="overlap"` - **(default for non-signed input)**:
Show overlap counts only.
3. `overlap_type="agreement"` - Show only **Agreement** (up/up and down/down),
and **Disagreement** (up/down and down/up).
4. `overlap_type="each"` - Show all combination of up and down directions.


### overlap_type="concordance"

```{r venndir_1, fig.alt="Venn diagram showing directional counts, summarized using 'concordance'."}
venndir(setlist_dir,
   font_cex=c(1.5, 1.5, 1),
   main='overlap_type="concordance"')
```

This approach is effective at conveying direction, without too many details.

### overlap_type="each"

```{r venndir_each, fig.alt="Venn diagram showing directional counts, summarized using 'each'."}
venndir(setlist_dir, overlap_type="each",
   font_cex=c(1.5, 1.5, 1),
   main='overlap_type="each"')
```

This option shows the count for each combination.

### overlap_type="agreement"


```{r venndir_agreement, fig.alt="Venn diagram showing directional counts, summarized using 'agreement'."}
venndir(setlist_dir,
   font_cex=c(1.5, 1.5, 1),
   overlap_type="agreement",
   main='overlap_type="agreement"')
```

This option is preferred for overall agreement, without all the details.

### overlap_type="overlap"

```{r venndir_overlap, fig.alt="Venn diagram showing directional counts, summarized using 'overlap'."}
venndir(setlist_dir,
   font_cex=c(1.5, 1.5, 1),
   overlap_type="overlap",
   main='overlap_type="overlap"')
```

Venn "Classic".


## Euler Direction

Add argument `proportional=TRUE` to
display a proportional Venn diagram (Euler diagram),
which uses the excellent `eulerr` R package.

```{r venndir_each_p, fig.alt="Proportional Euler diagram showing directional counts."}
vo <- venndir(setlist_dir,
   proportional=TRUE,
   font_cex=c(1.3, 0.9, 0.7))
```

Labeling is a challenge with proportional Venn diagrams,
venndir has several options to help optimize, and customize
label placement. (Bookdown docs forthcoming.)

For very small overlap regions, `inside_percent_threshold=5` will
move labels outside when the area is less than 5% of the total.

```{r venndir_overlap_p, fig.alt="Proportional Euler diagram with modified count labels."}
venndir(setlist_dir,
   proportional=TRUE,
   label_style="lite box",
   inside_percent_threshold=5,
   font_cex=c(1.3, 1))
```

The signed labels are arranged in "wide" format, but
"tall" may be more visually appealing.

```{r venndir_overlap_tall_p, fig.alt="Proportional Euler diagram with count labels arranged in 'tall' format."}
venndir(setlist_dir,
   proportional=TRUE,
   template="tall",
   label_style="lite box",
   font_cex=c(1.3, 1))
```

## Customizing the Venn diagram

Labels can be customized with shading, border, and placed inside
or outside the Venn area.

### Label styles

The argument `label_style` can be used to customize the label:

* `label_style="lite"` - adds lite shading behind each label
* `label_style="shaded"` - adds partially transparent shading
* `label_style="fill"` - adds solid colored shading
* `label_style="box"` - adds a small outline box around the label

Multiple terms can be combined, for example to add shading and a box:

* `label_style="shaded box"` - adds colored shading and a box outline

### Label position

Argument `show_labels` is used to position labels.
Each letter defines a type of label, and UPPERCASE or lowercase
indicates where to place the label.

* UPPERCASE = outside the Venn diagram
* lowercase = inside the Venn diagram

The letters:

* **N** = the set name
* **C** = the overlap count
* **S** = the signed overlap count(s)
* **i** = the overlapping items

Guidance:

* The default: `show_labels="Ncs"` will show _N_ame outside, _c_ounts inside.
When _s_igned labels are shown, they also appear inside.
Signed labels are not shown when `overlap_type="overlap"`.
* Display all labels inside: `show_labels="ncs"`
* It works best to have the _c_ounts and _s_igned counts together,
usually inside.
* To display items, `show_labels="Ni"` is recommended, to show _N_ame
outside, and _i_tems inside. You can still use `show_items="sign item"`
so that each item label will include the direction.
* When displaying items, the counts and signed counts are automatically
moved outside. (There isn't a great way to place item labels around
the count labels. Maybe in future.)


```{r label_preset_1, fig.alt="Venn diagram showing set label placement inside the Venn circles"}
setlist <- make_venn_test(1000, 3)
vo4 <- venndir(setlist,
   show_labels="ncs",
   main="Set labels are placed inside.",
   inside_percent_threshold=0)
```

Hide line segments with `show_segments=FALSE`

```{r label_preset_1l, fig.alt="Venn diagram with set labels outside the Venn circles, and line segments are hidden."}
vo4l <- venndir(setlist,
   show_labels="Ncs",
   show_segments=FALSE,
   main="Set labels are outside, with hidden line segments.",
   inside_percent_threshold=0)
```

## Highlights

Specific overlaps can be modified or highlighted, using either
`modify_venndir_overlap()` or `highlight_venndir_overlap()`.
Labels can be adjusted, moved, resized, etc.

```{r highlights, fig.alt="Venn diagram with a specific overlap highlighted with yellow fill color."}
vo4h <- highlight_venndir_overlap(vo4l,
   outerborder="red", outerborder.lwd=3,
   overlap_set="set_A&set_B")
render_venndir(vo4h, main="Highlight for **set_A&set_B**")
```

```{r highlights_2, fig.alt="Venn diagram with all overlaps involving set B highlighted with yellow fill color."}
vo4h <- highlight_venndir_overlap(vo4l,
   overlap_set=unique(grep("set_B", vo4l@jps@polygons$venn_name, value=TRUE)))
render_venndir(vo4h, main="All overlaps **set_B** and not **set_D**")
```

## Text Venn for the R Console

Text Venn diagrams are surprisingly convenient for remote server work!
(Output is colorized, not colorized in RMarkdown.)

The first example is the basic Venn overlap, without direction.

```{r textvenn_1, results='asis'}
setlist <- make_venn_test(1000, 3, do_signed=TRUE)
names(setlist) <- gsub("set_", "", names(setlist));
textvenn(setlist, overlap_type="overlap", htmlOut=TRUE)
```

But of course direction is helpful, so here it is with the
default `overlap_type="concordance"` (below)

```{r textvenn_2, results='asis'}
textvenn(setlist, overlap_type="concordance", htmlOut=TRUE)
```

Use `unicode=FALSE` if the console font does not support unicode arrows.

```{r textvenn_3, results='asis'}
textvenn(setlist, overlap_type="concordance", unicode=FALSE, htmlOut=TRUE)
```



## Nudge Venn circles

Sometimes when making a proportional Euler diagram, the
overlaps cannot all be displayed. Consider this example:

```{r nudge_1, fig.alt="Euler diagram showing overlaps and count labels. One overlap is not represented because the Euler circles did not represent it."}
overlaps <- c(set_A=187, set_B=146, set_C=499,
   `set_A&set_B`=1,
   `set_A&set_C`=181,
   `set_B&set_C`=219,
   `set_A&set_B&set_C`=20);
# convert to setlist
setlist_o <- counts2setlist(overlaps)

vn <- venndir(setlist_o,
   expand_fraction=0.15,
   proportional=TRUE,
   outerborder="white", outerborder.lwd=2, innerborder=NA,
   font_cex=1.4,
   set_colors=c("firebrick2", "dodgerblue", "#BBBBBB"))
```

The overlap `set_A&set_B` has 1 count, but is not displayed.

Use argument `circle_nudge` to move a specific Venn circle using
x,y coordinates.

```{r nudge_2, fig.alt="Euler diagram after nudging one of the circles so that the missing overlap would be displayed."}
# define the nudge
circle_nudge <- list(
   set_A=c(1, 0),
   set_B=c(-1, 0))
# apply the nudge
vo_nudge <- venndir(setlist_o,
   font_cex=1.4,
   proportional=TRUE, inside_percent_threshold=2,
   outerborder="white", outerborder.lwd=2, innerborder=NA,
   circle_nudge=circle_nudge,
   set_colors=c("firebrick2", "dodgerblue", "#BBBBBB"))
```


## Item labels

When looking at a Venn diagram,
a common *very next question* is  
"**What items are shared?**"

Add `"i"` to the `show_labels` argument, for example  
`show_labels="Ni"`

```{r vennitems_1, fig.alt="Venn diagram with item labels inside the Venn overlap regions."}
setlist <- make_venn_test(33, 3, do_signed=TRUE, item_prefix="item");
vo <- venndir(setlist,
   item_cex=0.8,
   show_labels="Ni")
```

You can show both the sign and the item name, or just one.  
Argument `show_items="sign"` will display only the direction.

```{r vennitems_1p, fig.alt="Venn diagram with directional arrows inside the Venn diagram overlap regions."}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
   fontfamily="Arial",
   show_labels="Ni",
   item_cex=1.5,
   show_items="sign",
   proportional=TRUE);
```

The sign is an interesting visual summary when there are too
many labels to display otherwise.

```{r vennitems_2, fig.alt="Venn diagram with a large number of directional arrows inside the Venn overlap regions."}
setlist <- make_venn_test(850, 3, do_signed=TRUE);
venndir(setlist,
   fontfamily="Arial",
   show_labels="Ni",
   show_items="sign",
   item_cex=1.5,
   item_buffer=0,
   max_items=10000);
```

## Venn Memes

More examples will be in the bookdown docs, and some fun ones
are in `venn_meme()`.  
Here is a simple example:

```{r venn_meme_bix, fig.alt="Venn meme showing three Venn circles, Stats, Computer Science, and Biology, as they relate to Bioinformatics as the intersection of all three."}
bix <- list(
   s="Stats",
   cs="Computer<br>Science",
   b="Biology",
   `s&b`="Biostatistics",
   `s&cs`="Data<br>Science",
   `cs&b`="Computational<br>Biology",
   `s&cs&b`="Bioinformatics")
venn_meme(bix,
   outerborder.lwd=0,
   innerborder="white")
```

You can provide a list of vectors, for example this fun example
comparing two cars:

```{r venn_meme, fig.alt="Venn meme comparing attributes of two cars, with labels shown in unique and shared sections of the Venn diagram."}
meme_list <- list(
   `Car A`=c("leather seats", "no sunroof", "SUV", "20mpg"),
   `Car B`=c("cloth seats", "sedan", "sunroof", "40mpg"),
   `Car A&Car B`=c("power locks", "power windows", "trunk storage"))
venn_meme(meme_list,
   font_cex=c(2, 1, 1),
   outerborder.lwd=0, innerborder.lwd=0,
   proportional=TRUE,
   # outerborder="white", outerborder.lwd=1, innerborder=NA,
   set_colors=c("orange", "firebrick3"),
   # poly_alpha=0.9,
   # item_buffer=-0.4,
   show_labels="Ni",
   show_segments=FALSE)
```


