---
output:
  github_document:
    fig_width: 5
    fig_height: 3
    df_print: kable
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE,
  message=FALSE,
  comment = "#>",
  fig.height=8,
  fig.width=8,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(knitr.table.format='markdown')
```

# venndir

<!-- badges: start -->
<!-- badges: end -->

The goal of venndir is to enable directional Venn overlap
analysis and visualization.


## Installation

The development version of venndir can be installed with:

``` r
# install.packages("remotes")
remotes::install_github("jmw86069/venndir");
```


## Brief Overview of venndir

The `venndir` package includes a utility function `make_venn_test()`
to create test data.

```{r venn_1}
library(venndir)
setlist <- make_venn_test(100, 3)
venndir(setlist)
```

The more interesting case uses directionality.

Note that `make_venn_tests()` can simulate concordance,
and the default is `concordance=0.5`. Concordance is defined
by `(agree - disagree) / (n)`, where `concordance=0` means
there are the same number that agree as disagree in direction,
and `concordance=1` means every element agrees in direction.

```{r venndir_1}
setlist <- make_venn_test(1000, 3, do_signed=TRUE)
venndir(setlist)
```

There are a few ways to represent concordance... sometimes you
want to see up-up, and down-down, and everything else is "mixed".
For that, use the default `overlap_type="concordance"` (shown above.)

Another option is to show "agreement" and "mixed", and for this
case use `overlap_type="agreement"` (below).

```{r venndir_agreement}
venndir(setlist, overlap_type="agreement")
```

The last option is to show the counts for each direction,
by `overlap_type="each"` (below).

```{r venndir_each}
venndir(setlist, overlap_type="each")
```

## Proportional directional Venn

You can even make a proportional Venn diagram, which uses the
really nice `eulerr` R package.

```{r venndir_each_p}
venndir(setlist, proportional=TRUE)
```

At this point, labeling is an issue, and the best current
remedy is manual adjustment. (In future, non-overlapping labels
will be available.)

The output of `venndir()` contains a `list` with
the polygon coordinates, and the label coordinates, so you
can manually adjust as needed.

Also, I really like having a function named `render_venndir()`.

```{r venndir_each_p2}
venndir_output <- venndir(setlist, proportional=TRUE, do_plot=FALSE);

# get label coordinate data
label_df <- venndir_output$label_df;

# get polygon data
venn_spdf <- venndir_output$venn_spdf;

# adjust set_A&set_C
ac_rows <- which(label_df$overlap_set %in% "set_A&set_C");
label_df <- venndir_output$label_df;
label_df[ac_rows,"x"] <- label_df[ac_rows,"x"] - 4.2;
label_df[ac_rows,"y"] <- label_df[ac_rows,"y"] + 1.3;

# be fancy and add color background to the label
label_df[ac_rows,"fill"] <- jamba::alpha2col(data.frame(venn_spdf)["set_A&set_C","color"], alpha=0.5);
label_df[ac_rows,"border"] <- data.frame(venn_spdf)["set_A&set_C","color"];

render_venndir(venn_spdf=venn_spdf, label_df=label_df, font_cex=1.3)

lines(x=label_df[ac_rows[1],"x"] + c(1.5, 4.2),
   y=label_df[ac_rows[1],"y"] + c(-0.4, -1.3))
```



## Text Venn?

There is a text Venn diagram, surprisingly useful for quickly
checking overlaps and direction.

The first example is the basic Venn overlap, without direction.

```{r textvenn_1}
textvenn(setlist, overlap_type="overlap")
```

But of course direction is helpful, so here it is with the
default `overlap_type="concordance"` (below)

```{r textvenn_2}
textvenn(setlist, overlap_type="concordance")
```

Not all consoles can display Unicode arrows, so you can use ASCII
output only with `unicode=FALSE`:

```{r textvenn_3}
textvenn(setlist, overlap_type="concordance", unicode=FALSE)
```

Sorry, no proportional text Venn diagrams (yet)!


## Nudge Venn circles

Another driving reason for this package is that sometimes
proportional Venn (Euler) diagrams fail to produce circles
that show all the overlaps. While sometimes it is
mathematically impossible, other times are just... puzzling.

See below, the overlap `set_A&set_B` has 1 count, but is not
displayed.

```{r nudge_1}
overlaps <- c(set_A=187, set_B=146, set_C=499,
   `set_A&set_B`=1,
   `set_A&set_C`=181,
   `set_B&set_C`=219,
   `set_A&set_B&set_C`=20);
# convert to setlist
setlist_o <- counts2setlist(overlaps)

venndir(setlist_o,
   proportional=TRUE,
   set_colors=c("red", "blue", "#9999AA"))
```

The argument `circle_nudge` lets you nudge (move) a Venn circle
given x,y coordinates.

```{r nudge_2}
venndir_output <- venndir(setlist_o,
   font_cex=2,
   proportional=TRUE,
   circle_nudge=list(set_A=c(1, 0), set_B=c(-1, 0)),
   set_colors=c("red", "blue", "#9999AA"))
```


## Item labels

An optional but useful feature is to include item labels
inside the Venn diagram. It helps answer the question,
"What are those shared items?" (In my experience, that's
a very early question.)

This step can also include the
directional sign, which helps indicate which items
are shared, and whether they have the same direction.
In cases with too many labels to display, it is sometimes
still useful to display the sign, as a visual cue for the
proportion of shared or discordant signs.


```{r vennitems_1}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist, display_items="sign item");
```

Interestingly, the density of labels gives some indication
of the relative overlaps.

The same plot using proportional circles makes the
label density effectively uniform:


```{r vennitems_1p}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
  display_items="item",
  proportional=TRUE);
```

With too many features to label, it's still interesting to indicate
the sign.

```{r vennitems_2}
setlist <- make_venn_test(10000, 3, do_signed=TRUE);
venndir(setlist, display_items="sign");
```

Again, proportional Venn circles effectively makes the density
uniform.

```{r vennitems_2p}
venndir(setlist,
  overlap_type="each",
  display_items="sign",
  proportional=TRUE);
```
