---
output:
  github_document:
    fig_width: 5
    fig_height: 3
    df_print: kable
    dev: ragg_png
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE,
  message=FALSE,
  comment = "#>",
  fig.height=7,
  fig.width=8,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(knitr.table.format='markdown')
```

# venndir

<!-- badges: start -->
<!-- badges: end -->

The `venndir` package provides Venn directional
diagrams, that can optionally display item labels
inside the overlap regions.

The `pkgdown` reference: [jmw86069.github.io/venndir](https://jmw86069.github.io/venndir)


## Installation

The development version of venndir can be installed with:

``` r
# install.packages("remotes")
remotes::install_github("jmw86069/venndir");
```

## Why Venndir?

Biological data often involves components, for example genes, and
the direction of change, for example increase or decrease. It is not
enough to compare genes affected by two experiments, it is critical
to compare the direction of change. How else can we discriminate
the effects of disease from the effects of a cure?

```{r venn_intro}
# silence the warnings
options("warn"=-1)

library(venndir)

vo <- venndir(make_venn_test(100, 2, do_signed=TRUE),
   proportional=TRUE,
   overlap_type="each", font_cex=1.2)

```

## Features of venndir

The core is `venndir()` which takes a `setlist` as
input and produces a Venn diagram. When the `setlist`
contains directionality (sign), the directional
overlaps are also displayed.

To illustrate the point, `make_venn_test()` is used
to create test `setlist` data.

```{r venn_setlist}
setlist <- make_venn_test(100, 3)
setlist
```

A `setlist` is a list of vectors. The names of the list,
one for each vector, are the set names. Each vector contains
items which are the subject of the Venn overlaps.

Given a `setlist`, you can create a Venn diagram with `venndir()`:

```{r venn_1}
venndir(setlist)
```

You can make a proportional Venn diagram,
also known as a Euler diagram. Add argument `proportional=TRUE`.

```{r venn_1e}
vo <- venndir(setlist,
   proportional=TRUE)
```


## Venn Direction

The namesake of this package is Venn with directionality!

Test with: `make_venn_test(..., do_signed=TRUE)`.
Each item is associated with a numerical direction:

* `+1` for up
* `-1` for down

Each vector in `setlist_dir` is a named vector,
whose **names** are the items, and whose **values** are
the direction, with `+1` or `-1`. Take a look.

```{r venn_setlist_signed}
setlist_dir <- make_venn_test(100, 3, do_signed=TRUE)
setlist_dir
```


For biological data, **direction is important and relevant**.
Up- or down-regulation might be the difference between disease
or treatment.

> Note `make_venn_tests()` can simulate concordance,
and the default is `concordance=0.5`. Concordance is a
measure of how frequently two directions are the same, and
is defined `(agree - disagree) / (n)`. Thus, `concordance=0` means
there are the same number that agree as disagree in direction,
and `concordance=1` means every element agrees in direction.

There are different ways to summarize directional overlaps:

* `overlap_type="concordance"` - **(default for signed input)**:
Concordant up \u2191, down \u2193, and discordant X.
* `overlap_type="overlap"` - **(default for basic input)**:
Overlap counts only.
* `overlap_type="agreement"` - Agreement and disagreement.
* `overlap_type="each"` - Counts for each combination of directions.


### overlap_type="concordance"

```{r venndir_1}
venndir(setlist_dir, font_cex=c(1.5, 1.5, 1))
```

This approach is effective at conveying direction, without too many details.

### overlap_type="each"

```{r venndir_each}
venndir(setlist_dir, overlap_type="each")
```

This option shows the count for each combination.

### overlap_type="agreement"


```{r venndir_agreement}
venndir(setlist_dir, overlap_type="agreement")
```

This option is preferred for overall agreement, without all the details.

### overlap_type="overlap"

```{r venndir_overlap}
venndir(setlist_dir, overlap_type="overlap")
```

Venn "Classic".


## Proportional Venn Direction

Add argument `proportional=TRUE` to
display a proportional Venn diagram (Euler diagram),
which uses the excellent `eulerr` R package.

```{r venndir_each_p}
vo <- venndir(setlist_dir,
   proportional=TRUE,
   font_cex=c(1.3, 0.9, 0.7))
```

Labeling is a challenge with proportional Venn diagrams,
venndir has several options to help optimize, and customize
label placement. (Bookdown docs forthcoming.)

For very small overlap regions, `inside_percent_threshold=5` will
move labels outside when the area is less than 5% of the total.

```{r venndir_overlap_p}
venndir(setlist_dir,
   proportional=TRUE,
   label_style="lite box",
   inside_percent_threshold=5,
   font_cex=c(1.3, 1))
```


## Customizing the Venn diagram

Labels can be customized with shading, border, and placed inside
or outside the Venn area.

### Label styles

The argument `label_style` can be used to customize the label:

* `label_style="lite"` - adds lite shading behind each label
* `label_style="shaded"` - adds partially transparent shading
* `label_style="fill"` - adds solid colored shading
* `label_style="box"` - adds a small outline box around the label

Multiple terms can be combined, for example to add shading and a box:

* `label_style="shaded box"` - adds colored shading and a box outline

### Label position

Argument `show_labels` is used to position labels.
Each letter defines a type of label, and UPPERCASE or lowercase
indicates where to place the label.

* UPPERCASE = outside the Venn diagram
* lowercase = inside the Venn diagram

The letters:

* N = the set name
* C = the overlap count
* S = the signed overlap count(s)
* i = the overlapping items

Guidance:

* The default: `show_labels="Ncs"` will show _N_ame outside, _c_ounts inside.
When _s_igned labels are shown, they also appear inside.
Signed labels are not shown when `overlap_type="overlap"`.
* Display all labels inside: `show_labels="ncs"`
* It works best to have the _c_ounts and _s_igned counts together,
usually inside.
* To display items, `show_labels="Ni"` is recommended, to show _N_ame
outside, and _i_tems inside. You can still use `show_items="sign item"`
so that each item label will include the direction.
* When displaying items, the counts and signed counts are automatically
moved outside. (There isn't a great way to place item labels around
the count labels. Maybe in future.)


```{r label_preset_1}
vo4 <- venndir(setlist,
   show_labels="ncs",
   inside_percent_threshold=0)
```

Hide line segments with `show_segments=FALSE`

```{r label_preset_1l}
vo4l <- venndir(setlist,
   show_labels="Ncs",
   show_segments=FALSE,
   inside_percent_threshold=0)
```

## Highlights

Specific overlaps can be modified or highlighted, using either
`modify_venndir_overlap()` or `highlight_venndir_overlap()`.
Labels can be adjusted, moved, resized, etc.

```{r highlights}
vo4h <- highlight_venndir_overlap(vo4l,
   outerborder="red", outerborder.lwd=3,
   overlap_set="set_A&set_B")
render_venndir(vo4h, main="Highlight for **set_A&set_B**")

vo4h <- highlight_venndir_overlap(vo4l,
   overlap_set=unique(grep("set_B", vo4l@jps@polygons$venn_name, value=TRUE)))
render_venndir(vo4h, main="All overlaps **set_B** and not **set_D**")

```

## Text Venn for the R Console

Text Venn diagrams are surprisingly convenient for remote server work!
(Output is colorized, not colorized in RMarkdown.)

The first example is the basic Venn overlap, without direction.

```{r textvenn_1, results='asis'}
# Options are used for the RMarkdown
# options("jam.htmlOut"=TRUE, "jam.comment"=FALSE)

setlist <- make_venn_test(1000, 3, do_signed=TRUE)
names(setlist) <- gsub("set_", "", names(setlist));
textvenn(setlist, overlap_type="overlap", htmlOut=TRUE)
```

But of course direction is helpful, so here it is with the
default `overlap_type="concordance"` (below)

```{r textvenn_2, results='asis'}
textvenn(setlist, overlap_type="concordance", htmlOut=TRUE)
```

Use `unicode=FALSE` if the console font does not support unicode arrows.

```{r textvenn_3, results='asis'}
textvenn(setlist, overlap_type="concordance", unicode=FALSE, htmlOut=TRUE)
```



## Nudge Venn circles

The overlap `set_A&set_B` has 1 count, but is not displayed.

```{r nudge_1}
overlaps <- c(set_A=187, set_B=146, set_C=499,
   `set_A&set_B`=1,
   `set_A&set_C`=181,
   `set_B&set_C`=219,
   `set_A&set_B&set_C`=20);
# convert to setlist
setlist_o <- counts2setlist(overlaps)

vn <- venndir(setlist_o,
   expand_fraction=0.15,
   proportional=TRUE,
   outerborder="white", outerborder.lwd=2, innerborder=NA,
   font_cex=1.4,
   set_colors=c("firebrick2", "dodgerblue", "#BBBBBB"))
```

Use argument `circle_nudge` to move a Venn circle using
x,y coordinates.

```{r nudge_2}
vo_nudge <- venndir(setlist_o,
   font_cex=1.4,
   proportional=TRUE, inside_percent_threshold=2,
   outerborder="white", outerborder.lwd=2, innerborder=NA,
   circle_nudge=list(set_A=c(1, 0), set_B=c(-1, 0)),
   set_colors=c("firebrick2", "dodgerblue", "#BBBBBB"))
```


## Item labels

Adding item labels can be convenient for the most common next
question: ***"What are those?"***

Add "i" to the `show_labels` argument, for example
change `show_labels="Nc"` to `show_labels="Ni"`.

Optionally adjust `show_items="sign item"` to include the sign, the item,
or both (default).


```{r vennitems_1}
setlist <- make_venn_test(33, 3, do_signed=TRUE, item_prefix="item");
vo <- venndir(setlist,
   item_cex=0.8,
   show_segments=FALSE,
   poly_alpha=0.4,
   sign_count_delim="",
   outerborder="white", outerborder.lwd=1, innerborder=NA,
   show_labels="Ni",
   # fontfamily="Arial",
   show_items="sign item");
```

Note that `show_items="sign"` displays only the directional arrow,
used in the proportional Euler diagram.

```{r vennitems_1p}
setlist <- make_venn_test(100, 3, do_signed=TRUE);
venndir(setlist,
   poly_alpha=0.3,
   show_labels="Ni",
   item_cex=1.5,
   show_items="sign",
   proportional=TRUE);
```

The sign is an interesting visual summary when there are too
many labels to display otherwise.

```{r vennitems_2}
setlist <- make_venn_test(850, 3, do_signed=TRUE);
venndir(setlist,
   poly_alpha=0.3,
   # rotate_degrees=60,
   outerborder="white", outerborder.lwd=1, innerborder=NA,
   show_labels="Ni",
   show_items="sign",
   item_cex=1.5,
   item_degrees=15,
   item_buffer=-0.05,
   show_segments=FALSE,
   combine_size=FALSE,
   max_items=10000);
```

## Venn Memes

More examples will be in the bookdown docs, and some fun ones
are in `venn_meme()`.
Here is a simple example:

```{r venn_meme}
meme_list <- list(`Car A`=c("leather seats", "no sunroof", "SUV", "20mpg"),
   `Car B`=c("cloth seats", "sedan", "sunroof", "40mpg"),
   `Car A&Car B`=c("power locks", "power windows", "trunk storage"))
venn_meme(meme_list,
   font_cex=c(2, 1, 1),
   outerborder.lwd=0, innerborder.lwd=0,
   proportional=TRUE,
   # outerborder="white", outerborder.lwd=1, innerborder=NA,
   set_colors=c("orange", "firebrick3"),
   poly_alpha=0.9,
   item_buffer=-0.4, show_labels="Ni", show_segments=FALSE)
```


